\chapter{Accelerate framework}

\section{ARM architecture/NEON}

\subsection{ARMv7}
\begin{itemize}
\item Used in iphone 3Gs, ipad, iphone 4, ipod 3rd generation touch. 
\item SIMD unit called NEON
\item eg. the function 
\begin{verbatim}
vmul.f32 qo, q1, q2
\end{verbatim}
will multiply four 32-bit floats from q2 with four 32-bit floats from q1 and store the result in q0.
\item The point is that this is done {\bf simulataneously}.
\item NEON does not use double precision units.
\item NEON may use more power but, since the instructions will take less time, overall they may be more efficient.
\end{itemize}

\section{What can be done in the accelerate framework}

\begin{itemize}
\item New to iphone 4.0.
\item Consists of many libraries, only three of which are currently available in iphone os 4.0. These are vDSP, LAPACK and BLAS.
\item vDSP (Digital Signal Processing library). 
\item eg. Dot product
\begin{verbatim}
#include <Accelerate/Accelerate.h>

float dotProduct;
vDSP_dotpr(a, 1, b, 1, &dotProduct, n);  // take in pointers to two arrays 
\end{verbatim}
\item This code is 8x faster than simple for loop. We also use approx. 1/4 of the energy.
\item Important to note that this can be used on different hardware.
\end{itemize}

\subsection{LAPACK}
\begin{itemize}
\item Been around a long time.
\item High level linear algebra.
\item Matrix factorisations.
\item Eigenvalues/vectors
\item Single/Double precision
\item Real/complex data types
\item Different matrix types
\item Because it comes from Fortran, it is laid out in column major order
\begin{verbatim}
float b[n];
b[0]=5.f; b[1]  = 5.f; b[2] = 4.f; b[3] = -2.f;

__CLPK_integer nrhs = 1;
__CLPK_integer ipiv[n];
__CLPK_integer info;

sgesv_(&n, &nrhs, A, &n, ipiv, b, &n, &info);
printf(``The solution is: (\%f, \%f, \%f, \%f)\n'', b[0],b[1], b[2], b[3]);

\end{verbatim}
\item when we compile this we must link against the accelerate framework eg.
\begin{verbatim}
gcc lapack.c -framework Accelerate -std=c99 -o lapackExample
./lapackExample
\end{verbatim}
\item The -std=c99 flag specifies we are using the 1999 standard of C.
\end{itemize}

\subsection{BLAS}

\begin{itemize}
\item Low level stuff
\item eg. Dot product, vector products
\item Matrix-vector products
\item Matrix multiplies
\item Supports for both row and column major order
\item Can use dense and triangular matrices
\begin{verbatim}
#include <Accelerate/Accelerate.h>
#include <stdio.h>
int main()
{
float A[2][2] = {{1.f, 1.f}, 
		        {0.f, 1.f}};
float B[2][2] = {{1.f, 2.f}, 
		        {3.f, 4.f}};	
float C[2][2];

cblas_sgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans,
    		    2, 2, 2, 1.f, A, 2, B, 2, 0.f, C, 2);	
     		    
printf("%f, %f, %f, %f\n", C[0][0], C[0][1], C[1][2], C[2][2]);

return(0);
}        	        	
\end{verbatim}
\end{itemize}











